Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-20 src/tex/ms.tex
[showyourwork] overleaf sync

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to the large datasets produced by modern stellar surveys.}
  % aims heading (mandatory)
   {This study aims to leverage simulation-based inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances. By focusing on the high-mass IMF slope and SN\,Ia normalization, we aim to demonstrate the advantages of SBI over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combined a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train an SBI pipeline. Mock datasets were generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, and additional tests were conducted on data from a Milky Way-like galaxy taken from a hydrodynamical simulation. SBI results were benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {Our SBI approach achieves a $\sim1,000\times$ ($\sim72,000\times$) speed-up compared to HMC, reducing inference runtime from $\sim40$ hours for 200 stars to mere minutes for thousands of stars. Inference on $1,000$ stars yielded precise estimates for the IMF slope ($\alpha_{\rm IMF} = -2.296 \pm 0.009$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = -2.889 \pm 0.010$), deviating by less than 0.04\% from the ground truth. \textcolor{blue}{SBI also demonstrated greater robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation (not with the new model :( )}.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in galactic chemical enrichment studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys, such as 4MOST and WEAVE, facilitating deeper insights into the chemical and dynamical evolution of galaxies.}

>>>>

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to large datasets of modern stellar surveys.}
  % aims heading (mandatory)
   {We leverage simulation-based-inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances and demonstrate its the advantages over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combine a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train our SBI pipeline. Mock datasets are generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, with additional tests conducted on data from a simulated Milky Way-like galaxy. SBI results are benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {SBI achieves a $\sim1,000\times$ speed-up compared to HMC, reducing inference runtime from $\gtrsim40$ hours to mere minutes for thousands of stars. Inference on $1,000$ stars yields precise estimates for the IMF slope ($\alpha_{\rm IMF} = -2.296 \pm 0.009$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = -2.889 \pm 0.010$), deviating less than 0.04\% from the ground truth. SBI also demonstrates greater robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in GCE studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys facilitating deeper insights into the chemical and dynamical evolution of galaxies.}
